Density-dependent selection in evolutionary genetics: a lottery model of Grime's triangle

### 2017-01-10 manuscript.tex
latest

In the classic lottery model \citep{chesson_1981}, unoccupied territories are saturated with propagules from every genotype i.e. the mean territorial propagule density $l_i=m_i/U$ is much larger than one for all genotypes. The law of large numbers then implies that the composition of propagules in each territory will not deviate appreciably from the mean composition $l_1,l_2,\ldots,l_G$ ($G$ is the number of genotypes present), and so the probability that genotype $i$ wins any particular unoccupied territory is approximately $c_i l_i/\sum_j c_j l_j$. Since $T$ is large (by assumption), genotype $i$ wins approximately $c_i l_i/\sum_j c_j l_j$ of the $U$ available territories, and so the increase in $n_i$ due to territorial acquisition in the classic lottery model is given by
\begin{equation}
\Delta_+ n_i(t)=U(t)\frac{c_i l_i}{\sum_j c_j l_j}= b_i n_i\frac{1}{L}\frac{c_i}{\overline{c}}, \label{eq:lottery}
\end{equation}
where $\overline{c}=\sum_j c_j m_j/M$ is the mean propagule competitive ability for a randomly selected propagule, $L=M/U$ is the total propagule density and $M=\sum_j m_j$ is the total number of propagules.

Unlike the classic lottery model, we do not restrict ourselves to high propagule densities. In our extension to arbitrary densities, we expect that $p_1(x_1)\ldots p_G(x_G)$ of the $U$ unoccupied territories will have the propagule composition $x_1,\ldots,x_G$. Within these $U(t) p_1(x_1)\ldots p_G(x_G)$ territories,

>>>>

In the classic lottery model \citep{chesson_1981}, unoccupied territories are assumed to be saturated with propagules from every genotype i.e. the mean propagule density $l_i=m_i/U$ is much larger than one for all genotypes. From the law of large numbers, the composition of propagules in each territory will then not deviate appreciably from the mean composition $l_1,l_2,\ldots,l_G$ ($G$ is the number of genotypes present), and so the probability that genotype $i$ wins any particular unoccupied territory is approximately $c_i l_i/\sum_j c_j l_j$. Thus, genotype $i$ is expected to win $c_i l_i/\sum_j c_j l_j$ of the $U$ available territories, and deviations from this expected outcome are small (since $T$ is large by assumption). Consequently, the increase in $n_i$ due to territorial acquisition in the classic lottery model is given by
\begin{equation}
\Delta_+ n_i(t)=\frac{c_i l_i}{\sum_j c_j l_j}U(t)=b_i n_i\frac{1}{L}\frac{c_i}{\overline{c}}, \label{eq:lottery}
\end{equation}
where $\overline{c}=\sum_j c_j m_j/M$ is the mean propagule competitive ability for a randomly selected propagule, $L=M/U$ is the total propagule density and $M=\sum_j m_j$ is the total number of propagules.

In our extension of the classic lottery model, we do not restrict ourselves to high propagule densities, and the propagule composition in different territories will deviate appreciably from $l_1,l_2,\ldots,l_G$. We expect that a fraction $p_1(x_1)\ldots p_G(x_G)$ of the $U$ unoccupied territories will have the propagule composition $x_1,\ldots,x_G$. Genotype $i$ is expected to win win $c_i x_i/\sum_j c_j x_j$ of these $p_1(x_1)\ldots p_G(x_G) U(t)$ territories.
